---
title: "p8105_hw2_th3072"
author: "Tianyou He"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
```

# Problem 1
## part 1 `clean the data in pols-month.csv`
```{r}
pols = 
  read.csv("./hw2_data/pols-month.csv")|> 
  separate(mon, into = c("year", "month", "day"), convert = TRUE) |> 
  mutate(
    month = month.abb[as.numeric(month)], 
    president = recode (prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")
    ) |>
  select(-prez_dem, -prez_gop, -day) 
```

## part 2 `clean the data in snp.csv`
```{r}

snp =
  read.csv("./hw2_data/snp.csv")|>
  separate(date, into = c("month", "day", "year"), convert = TRUE)|>
  mutate(month = month.abb[as.numeric(month)], year = ifelse(year >=50, 1900+year, 2000+year))|>
  select(year, month, everything())|>
  arrange(year, month)|>
  select(-day)
```

## part 3 `tidy the unemployment data`
```{r}
unemployment = 
  read.csv("./hw2_data/unemployment.csv")|>
  rename(year = Year)|>
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment_rate"
  )
```

## part 4 `merge data`
```{r}
merged_data <- pols|> 
  left_join(snp, by = c("year","month"))|>
  left_join(unemployment, by = c("year", "month"))
```

*brief description*

For `pols` dataset, there are `r nrow(pols)` observation and the key variable here should be `president`, showing that `dem` or `gop`. The other variable also classified by `dem` and `gop`. The time for this dataset is from 1947 to 2015.

For `snp` dataset, there are `r nrow(snp)` observation and the key variable here should be `close`. The time for this dataset is from 1950 to 2015. 

For `unemployment` dataset, there are `r nrow(unemployment)` observation and the key variable here should be `unemployment rate`. The time for this dataset is from 1948 to 2015. 

After combine all these three datasets, there are `r nrow(merged_data)` observation and the key variable should be `president` and `unemployment rate`. For different president, they have different policy, so that the unemployment rate can show whether the policy is good or bad. The time for this combined dataset is from 1947 to 2015.

# Problem 2

## Mr. Trash Whell
```{r}
homes_per_ton <- 500/30 #define homes_per_ton for later home power calculation
```

```{r}
mr_trash_wheel = 
  read_excel("./hw2_data/202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel")|>
  rename(
    weight = `Weight (tons)`,
    volume = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = `Polystyrene`,
    cigarette_butts = `Cigarette Butts`,
    glass_bottles = `Glass Bottles`,
    plastic_bags = `Plastic Bags`,
    wrappers = `Wrappers`,
    sport_balls = `Sports Balls`,
    homes_powered = `Homes Powered*`
  )|>
  select(weight, volume, plastic_bottles,polystyrene, cigarette_butts, glass_bottles, plastic_bags, wrappers, sport_balls, homes_powered)|>
  mutate(homes_powered = weight * homes_per_ton)|>
  mutate( source = "mr_trash_wheel")

mr_trash_wheel <- mr_trash_wheel[-nrow(mr_trash_wheel),]
```

## Professor Trash Wheel
```{r}
professor_trash_wheel = 
  read_excel("./hw2_data/202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel")|>
  rename(
    weight = `Weight (tons)`,
    volume = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = `Polystyrene`,
    cigarette_butts = `Cigarette Butts`,
    glass_bottles = `Glass Bottles`,
    plastic_bags = `Plastic Bags`,
    wrappers = `Wrappers`,
    homes_powered = `Homes Powered*`
  )|>
  select(weight, volume, plastic_bottles,polystyrene, cigarette_butts, glass_bottles, plastic_bags, wrappers, homes_powered)|>
  mutate(homes_powered = weight * homes_per_ton)|>
  mutate(source = "professor_trash_wheel")

professor_trash_wheel <-professor_trash_wheel[-nrow(professor_trash_wheel),]

total_weight <- sum(professor_trash_wheel$weight)
```

## Gwynnda Trash Wheel
```{r}
Gwynnda_trash_wheel = 
  read_excel("./hw2_data/202309 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel")|>
  rename(
    weight = `Weight (tons)`,
    volume = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = `Polystyrene`,
    cigarette_butts = `Cigarette Butts`,
    plastic_bags = `Plastic Bags`,
    wrappers = `Wrappers`,
    homes_powered = `Homes Powered*`
  )|>
  select(weight, volume, plastic_bottles,polystyrene, cigarette_butts,plastic_bags, wrappers, homes_powered)|>
  mutate(homes_powered = weight * homes_per_ton)|>
  mutate( source = "Gwynnda_trash_wheel")

Gwynnda_trash_wheel <-Gwynnda_trash_wheel[-nrow(Gwynnda_trash_wheel),]
```

```{r}
gwynnda_cig = 
  read_excel("./hw2_data/202309 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel")|>
  rename(
  cigarette_butts = `Cigarette Butts`
  )|>
  select(Year, Month, cigarette_butts)

gwynnda_cig_filter = filter(gwynnda_cig, Year == "2021", Month == "July") #filter gwynnda cigarette butts on July, 2021

gwynnda_cig_sum = sum(gwynnda_cig_filter$cigarette_butts)
```

```{r}
all_data <- bind_rows( mr_trash_wheel, professor_trash_wheel, Gwynnda_trash_wheel) #combine all data
```
*brief description*

There are four dataset here: `mr_trash_wheel`, `professor_trash_wheel`, `Gwynnda_trash_wheel`, and `all_data`.
`r nrow(mr_trash_wheel)` observation for mr_trash_whell, `r nrow(professor_trash_wheel)` observation for professor_trash_wheel,and `r nrow(Gwynnda_trash_wheel)` observation for Gwynnda_trash_wheel.

For `all_data`, there are `r nrow(all_data)` observation and key variables for these four datasets are `weight` and `home_powered`. 

The total weight of trash collected by `professor_trash_wheel` here is `r total_weight` and the total number of cigarette butts collected by Gwynnda in July of 2021 is `r gwynnda_cig_sum`.


# Problem 3

## baseline data

```{r}
baseline_origin <-read.csv("./hw2_data/MCI_baseline.csv", skip = 1)
```

```{r}
baseline <-read.csv("./hw2_data/MCI_baseline.csv")|> #import MCI_baseline dataset
  rename(
  id = `X`,
  sex = `X1...Male..0...Female`,
  apoe4 = `X1...APOE4.carrier..0...APOE4.non.carrier`,
  age_at_study = `Age.at.the.study.baseline`,
  age_onset_MCI_and_missing = `Age.at.the.onset.of.MCI..missing.if.a.subject.remains.MCI.free.during.the.follow.up.period`,
  year_of_education = `Years.of.education`
) #rename the variable
baseline$sex <- ifelse(baseline$sex ==1, "male",
                       ifelse( baseline$sex == 0, "female", "Sex")) # change 1 and 0 to male and female
baseline$apoe4 <- ifelse(baseline$apoe4 == 1, "carrier",
                         ifelse(baseline$apoe4 == 0, "non-carrier", "apoe4")) # change 1 and 0 to carrier and non-carrier


baseline <- baseline|>
  filter(age_at_study < age_onset_MCI_and_missing| age_onset_MCI_and_missing == ".") # filter current age < age onset mci and ignore "."
```


```{r}
n_participants <- nrow(baseline) #number of participant 

only_mci <- baseline |>
  filter(age_at_study < age_onset_MCI_and_missing)
n_mci <- nrow(only_mci) # people have MCI

baseline$age_at_study <- as.numeric(baseline$age_at_study)
average_age <- mean(baseline$age_at_study, na.rm = TRUE) # find average baseline age 

women = filter(baseline, sex =="female")
women_carrier = filter(baseline,  sex == "female", apoe4 == "carrier")
women_apoe4 <- nrow(women_carrier)/nrow(women) # find the propotion of women carrier
```

## amyloid data
```{r}
amyloid <-read.csv("./hw2_data/mci_amyloid.csv")|>
  rename(
  id = `Study.ID`,
  time2 = `NA.`,
  time4 = `NA..1`,
  time6 = `NA..2`,
  time8 = `NA..3`,
  baseline = `Time..in.years..elapsed.since.the.study.baseline.to.the.visit.where.biomarker.Amyloid._.42.40.ratio.was.measured`
)

```


```{r}
in_baseline = anti_join(baseline, amyloid, by = "id")# in baseline not in amyloid

```

```{r}
in_amyloid = anti_join(amyloid, baseline, by = "id")# in amyloid not in baseline

in_amyloid <- in_amyloid[-1,]
```

```{r}
combined <- inner_join(baseline, amyloid, by = "id") # combine baseline and amyloid
write_csv(combined, "./hw2_data/ combined baseline and amyloid.csv") # save document
```


*brief discription *

There are `r nrow(baseline)` participant were recruited, and of these, there are `r nrow(only_mci)` people develop the MCI. 

The average baseline age is `r average_age`.

The proportion of women in the study are APOE4 carriers is `r women_apoe4`. 

Moreover, number people only in baseline but not in amyloid are `r nrow(in_baseline)` and number only in amyloid but not in baseline `r nrow(in_amyloid)`. After combining these two data, there are `r nrow(combined)` people in both baseline and amyloid dataset. This shows that there is a strong relationship between baseline and amyloid. 

